---
title: "crime classification EDA"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Import data and review names
```{r}
SCC = read.csv("train.csv", header = TRUE)
names(SCC)
```

# Clear difference in crimes by day of week; crimes peak on Friday see lowest levels on Sunday and rise from Mon-Wed.
```{r}
summary(SCC$DayOfWeek)
```

# There are some outliers in terms of Longitude.
```{r}
summary(SCC$X)
summary(SCC$Y)
```

# 67 Cases have a latitude of 90 and longitude of -120.5. This falls off of the lat/longitude map wherease other values fall directly in San Francisco. These cases should likely be removed.
```{r}
print(subset(SCC, Y > 40))
```

# Specific dates are a factor; top crime dates are at the start/end of the new year suggesting holidays should be explored.
```{r}
str(SCC$Dates)
summary(SCC$Dates)[0:20]
```

# Southern and Mission Districts have the most crime during this whole period; need to see whether trends have been changing in these areas.
```{r}
summary(SCC$PdDistrict)
```

# Top crime categories include Larceny/Theft, Other_Offenses, Non_Criminal. Some categories such as "Trea" (which has six cases of Trespessing) have few examples.
```{r}
summary(SCC$Category)
```

# "Other offfenses"" is a broad category, but driven by driving/traffic violations, and Evasion of Legal Requirements (Probation Violation, Parole Violation, Violation of Restraining Orders, etc.).  And my personal favorite "DANGER OF LEADING IMMORAL LIFE."

```{r}
SCC_OtherOffenses = subset(SCC, SCC$Category == "OTHER OFFENSES")
summary(SCC_OtherOffenses$Descript)[0:50]
```

# Larceny/Theft is a combination of Grand Larceny and Petty Theft.
```{r}
SCC_Larceny_Theft = subset(SCC, SCC$Category == "LARCENY/THEFT")
summary(SCC_Larceny_Theft$Descript)[0:20]
```

# Non-Criminal (one of the largest categories) is not related to crime and may cause noise in the analysis.
```{r}
SCC_Non_Criminal = subset(SCC, SCC$Category == "NON-CRIMINAL")
summary(SCC_Non_Criminal$Descript)[0:20]
```

# Assault is what you would expect, and also has several categories involving family/partner/child/elder abuse.
```{r}
SCC_Assault = subset(SCC, SCC$Category == "ASSAULT")
summary(SCC_Assault$Descript)[0:20]
```

# DRUG/NARCOTIC straightforward with the main distinction being possession vs. possession for sale
```{r}
SCC_DRUG_NARCOTIC = subset(SCC, SCC$Category == "DRUG/NARCOTIC")
summary(SCC_DRUG_NARCOTIC$Descript)[0:20]
```

# VANDALISM: 'Malicious Mischief" largest category is for vandalism of vehicles.
```{r}
SCC_VANDALISM= subset(SCC, SCC$Category == "VANDALISM")
summary(SCC_VANDALISM$Descript)[0:10]
```


# WARRANTS: not sure about this one but it seems like crimes have already been committed and we don't hvae details on those crimes. May duplicate with other crime stats.
```{r}
SCC_WARRANTS= subset(SCC, SCC$Category == "WARRANTS")
summary(SCC_WARRANTS$Descript)[0:10]
```

# SUSPICIOUS OCC:  It feels like crime is about to happen!
```{r}
SCC_SUSPICIOUS_OCC= subset(SCC, SCC$Category == "SUSPICIOUS OCC")
summary(SCC_SUSPICIOUS_OCC$Descript)[0:10]
```


# MISSING PERSON:  Only categories are Found Person (9317), Missing Juvenile (8620), Missing Adult (8052). 
```{r}
SCC_MISSING_PERSON= subset(SCC, SCC$Category == "MISSING PERSON")
summary(SCC_MISSING_PERSON$Descript)[0:10]
```

# ROBBERY: Fairly straightforward, usually on the street with either physical force, a knife or a gun.
```{r}
SCC_ROBBERY= subset(SCC, SCC$Category == "ROBBERY")
summary(SCC_ROBBERY$Descript)[0:20]
```

# "SECONDARY CODES" are opaque in title and descriptors.
```{r}
SCC_SECONDARY_CODES= subset(SCC, SCC$Category == "SECONDARY_CODES")
summary(SCC_SECONDARY_CODES$Descript)[0:10]
```

# Family Offenses - Desertion of a Child, Minor without proper care, etc.
```{r}
SCC_FAMILY_OFFENSES= subset(SCC, SCC$Category == "FAMILY OFFENSES")
summary(SCC_FAMILY_OFFENSES$Descript)[0:10]
```
